Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-19 paper.tex
Ideal scan architecture fix

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work. \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

@@

\begin{figure*}[!t]
  \centering
  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=7.5cm,keepaspectratio]{graphics/ScanThenPropagate.pdf}
      \caption{\emph{Scan-then-Propagate.}}
      \label{fig:scan-then-propagate}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/ReduceThenScan.pdf}
      \caption{\emph{Raking Reduce-then-Scan.}}
      \label{fig:reduce-then-scan}
    \end{subfigure}%
  }\\[1em]

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-then-Scan} represent two contrasting paradigms in scan design. By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for in-situ operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple's GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The Reduce-then-Scan threshold is the expected execution time of RTS at the same input size; runs to the left of this threshold indicate a performance uplift and vice versa. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure*}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We demonstrate that work-stealing is particularly well suited to \emph{Chained Scan}. By imposing blocking rates of up to 50\% to simulate unfair schedulers, we show that \emph{Decoupled Fallback} outperforms \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}~\cite{Merrill2016}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{speed-of-light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite \emph{starvation} and will never complete. In practice, it is easy to demonstrate this highly undesirable behavior.

@@

This work is guided by two main goals: \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near-SOL execution to the greatest extent possible allowed by the underlying hardware and programming model.

However, certain aspects fall outside the scope of this work: \textbf{Performance guarantees} are not universally attainable, as not all architectures offer atomic operations that are sufficiently fast or scheduling models that are sufficiently fair to attain SOL performance. \textbf{Subgroup divergence} is another challenge we acknowledge but do not address directly, as shading languages do not provide explicit divergence control like CUDA's subgroup masking mechanisms. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained Scan}.

@@

\begin{figure*}
  \centering
  \makebox[\textwidth][c]{
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=7.5cm,keepaspectratio]{graphics/ScanThenPropagate.pdf}
      \caption{\emph{Scan-then-Propagate.}}
      \label{fig:scan-then-propagate}
    \end{subfigure}
    \hfill
    \begin{subfigure}[c]{0.48\textwidth}
      \centering
      \includegraphics[height=6cm,keepaspectratio]{graphics/ReduceThenScan.pdf}
      \caption{\emph{Raking Reduce-then-Scan.}}
      \label{fig:reduce-then-scan}
    \end{subfigure}%
  }\\[1em]

@@

\subsection{Ideal Scan Architecture}
In terms of performance, an ideal global scan algorithm should minimize global memory movement, fully utilize the device’s memory bandwidth, and avoid costly global barriers. By these criteria alone, the \emph{Chained Scan} architecture emerges as the preferred solution; moreover, its single-kernel design offers compelling portability advantages across diverse hardware. By eliminating global barriers, \emph{Chained Scan} naturally conforms to the GPU’s oversubscription model, allowing occupancy control to be delegated the hardware's workgroup scheduler. In contrast, hierarchical approaches such as \emph{Reduce-then-Scan} require workgroup raking and fixed-occupancy persistent threads to bound the spine’s size, necessitating explicit kernel launches and manual tuning of launch configurations to maintain optimal occupancy.

Hardware-level considerations underscore the limitations of the fixed-launch approach. On GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—fixed launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns—or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations—may experience significant performance degradation.

Moreover, the fused nature of \emph{Chained Scan} enables in-situ compaction: algorithms requiring compaction can merge that work directly into the scan kernel, avoiding redundant writes or secondary compaction passes. This fusion benefits applications ranging from data deduplication to compression. However, \emph{Chained Scan} has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient—it risks indefinite starvation, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
 Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite[Section~2.11]{HSASysArch12} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their 2021 findings~\cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis~\cite{sorensen2018}.

@@

\subsection{Why is Reliance on Forward-Progress Guarantees a Portability Problem?}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---are eventually given execution and make progress toward termination, preventing indefinite starvation. In the most comprehensive 
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing (Figure~\ref{fig:dist}) shows that, at best, running \emph{Chained Scan} without FPG results in performance worse than \emph{Reduce-then-Scan}, and frequently triggers \emph{Timeout Device Recovery}\label{sec:tdr}\footnote{Operating systems handle unresponsive programs differently. On Windows, the Timeout Detection and Recovery (TDR) system resets the GPU after approximately 2 seconds. Other operating systems---such as macOS and Linux---employ their own timeout mechanisms with varying behaviors. For consistency, we adopt Windows’s 2-second limit as a reasonable expectation for maximum execution time.} (TDR), crashing the program.
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/decoupled_lookback.pdf}
    \caption{\emph{Decoupled Lookback.}}
    \label{fig:decoupled}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\linewidth}
    \includegraphics[width=\linewidth]{graphics/ours.pdf}
    \caption{\emph{Decoupled Fallback (ours).}}
    \label{fig:ours}
  \end{subfigure}
  \caption{Histogram plots, 500 runs of Inclusive Prefix Sum, $\mathbf{2^{25}}$ 32-bit elements, on an FPG-lacking device, Apple M1 Max. Bar heights ($y$ axis) indicate the number of runs that completed in the execution time indicated on the $x$ axis. The Reduce-then-Scan threshold is the expected execution time of RTS at the same input size; runs to the left of this threshold indicate a performance uplift and vice versa. The TDR threshold is Windows's threshold for resetting the GPU\@; as noted in Section~\ref{sec:tdr}, MacOS handles this differently.}
  \label{fig:dist}
\end{figure*}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

\begin{figure*}
  \centering
  \includegraphics[width=.95\textwidth]{graphics/DecoupledFallback.pdf}
  \caption{\emph{Decoupled Fallback}. For brevity, the chained propagation phase is condensed, see \ref{fig:decoupled-lookback} for the full behavior.}
  \label{fig:decoupled-fallback}
\end{figure*}
